---
title: "R Notebook"
output: html_notebook
author: apawlik
---

Required packages
```{r}
library(tidyverse)
library(lubridate)
library(dplyr)
library(xlsx)
library(tibble)
library(dataMaid)
options(stringsAsFactors = FALSE)
```




Read CSV into data frames (older data only)
```{r}
weather_78_89 <- read.csv("src/Daily_Weather_Raw_1978-1989.csv")
weather_90_16 <- read.csv("src/Daily_Weather_Raw_1990-2016.csv")
```

Remove unwanted columns from older data
```{r}
# remove unwanted columns from 90-16
weather_90_16 <- weather_90_16[1:13]
```

Fix data types of older data
```{r}
# put dates in standard format
weather_78_89["Date"] <- as.Date(weather_78_89[[ "Date" ]], format = "%m/%d/%y")
weather_90_16["UMB"] <- as.Date(weather_90_16[[ "UMB" ]], format = "%m/%d/%y")
```

Compile older data into single data frame
```{r}
# examine column names, determine how to align
names(weather_78_89)
names(weather_90_16)

# change column names
names(weather_90_16)[1] <- "Date"

# combine 2 old datasets vertically
weather_older <- rbind(weather_78_89, weather_90_16)

# glimpse(weather_older)
```




Read Excel files into data frames (newer data only)
```{r}
# 32 as of May 2020, excludes notes sheet
num_sheets <- 32

# column types for imported data
columnTypes <- c("Date")

# make a list of monthly sheets beginning with oldest 2 months (so month indices won't change)
l1 <- read.xlsx("src/Weather_Edit_2017-Present.xlsx", num_sheets, colClasses = columnTypes, endRow = 32)
l2 <- read.xlsx("src/Weather_Edit_2017-Present.xlsx", num_sheets - 1, colClasses = columnTypes, endRow = 32)
weather_17_pres <- list(l1, l2)
for (i in (num_sheets - 2):2) {  # exclude current month (incomplete data), so last month is last in list
  l <- read.xlsx("src/Weather_Edit_2017-Present.xlsx", i, colClasses = columnTypes, endRow = 32)
  weather_17_pres <- append(weather_17_pres, list(l))
}

num_months <- num_sheets - 1  # exclude current month (incomplete data)
```

Remove unwanted rows and columns from newer data
```{r}
# check that each has correct number of rows and columns before combining
num_cols <- 17  # last column in excel sheet is Q

# remove columns outside num_cols
for (i in 1:num_months) {
  # oct2018 (sheet 13) contains an extra column (13), remove it
  if (i == 13) {  
    weather_17_pres[[i]] <- weather_17_pres[[i]][c(1:12, 14:(num_cols + 1))]
  } else {
    weather_17_pres[[i]] <- weather_17_pres[[i]][1:num_cols]
  }
}

# remove rows beyond number of days in that month
for (i in 1:num_months) {
  n <- days_in_month(as.Date(weather_17_pres[[i]][10, 1], "%Y-%m-%d"))
  names(n) <- NULL  # convert from named int to just int
  weather_17_pres[[i]] <- weather_17_pres[[i]][1:n, ]
}
```

Fix data types of newer data
```{r}
# convert datetimes from Time column back to just times
for (i in 1:num_months) {
  weather_17_pres[[i]]$Time <- format(strptime(weather_17_pres[[i]]$Time, "%Y-%m-%d %H:%M:%S"), "%H:%M")
  # print(weather_17_pres[[i]][2])
}
```

Compile newer data into single data frame
```{r}
# combine elements of list vertically
weather_newer <- rbind(weather_17_pres[[1]], weather_17_pres[[2]])
for (i in 3:num_months) {
  weather_newer <- rbind(weather_newer, weather_17_pres[[i]])
}

# glimpse(weather_newer)
```




Combine older with newer into final data frame
```{r}
# examine column names, determine how to align
names(weather_older)
names(weather_newer)

# convert to tibble to use dplyr, trim whitespaces
weather_older_t <- as_tibble(lapply(weather_older, trimws))
weather_newer_t <- as_tibble(lapply(weather_newer, trimws))

# check if any rows rely on knots data over mph data
weather_newer_t %>% 
  filter(!is.na(Wind.Speed..knots.) && is.na(Wind.Speed..mph.))

# remove wind speed knots column from newer (redundant)
weather_newer_t <- weather_newer_t %>%
  select(-Wind.Speed..knots.)

# older data has 3 missing columns
weather_older_t <- weather_older_t %>%
  mutate(Time = NA, .after = 1) %>% 
  mutate(Evaporation.Pan = NA, .after = 13) %>% 
  mutate(Lake.Level = NA, .after = 14)

# move precipitation column in old data
weather_older_t <- weather_older_t %>% 
  relocate(RAIN.GAUGE, .after = 6)

# columns can be renamed to be shorter with details moved to metadata
weather_newer_t <- weather_newer_t %>% 
  rename(Max_Temperature = X24.Hr.Temp.Max) %>% 
  rename(Min_Temperature = X24.Hr.Temp.Min) %>% 
  rename(Current_Temperature = Temp.Current) %>% 
  rename(Precipitation_Belfort = X24.Hr.Precip..Belfort.) %>% 
  rename(Precipitation_ETI = X24.Hr.Precip..ETI.) %>% 
  rename(Snowfall = X24.Hr.Snow..in.) %>% 
  rename(Snow_On_Ground = Official.Snow.on.Ground..in.) %>% 
  rename(Wind_Direction = Wind.Direction) %>% 
  rename(Wind_Speed = Wind.Speed..mph.) %>% 
  rename(Sky_Cover = Sky.Cover.) %>% 
  rename(Cloud_Type = Cloud.Type) %>% 
  rename(Evaporation_Pan = Evaporation.Pan) %>% 
  rename(Lake_Level = Lake.Level) %>% 
  rename(Notes = Other.Observations.Notes)

weather_older_t <- weather_older_t %>% 
  rename(Max_Temperature = Max.Temp) %>% 
  rename(Min_Temperature = Min.Temp) %>% 
  rename(Current_Temperature = Observed.Temp) %>% 
  rename(Precipitation_Belfort = Rain.Melted..snow..etc) %>% 
  rename(Precipitation_ETI = RAIN.GAUGE) %>% 
  rename(Snowfall = Snow..ice.pellets..hail) %>% 
  rename(Snow_On_Ground = Snow.ice.pellets..hail..ice.on.ground) %>% 
  rename(Wind_Direction = Wind.Direction) %>% 
  rename(Wind_Speed = Wind.Speed.MPH) %>%
  rename(Sky_Cover = Sky.Cover) %>% 
  rename(Cloud_Type = Cloud.Type) %>% 
  rename(Evaporation_Pan = Evaporation.Pan) %>% 
  rename(Lake_Level = Lake.Level)
  
# check
names(weather_older_t)
names(weather_newer_t)

# combine
weather <- rbind(weather_older_t, weather_newer_t)
glimpse(weather)
```

Check missing value codes
```{r}
# make sure no dates are missing
weather %>% filter(is.na(Date))
weather %>% filter(Date == "")

# make sure all missing values are NA
lapply(weather, identifyMissing)

# fill empty cells with NA
weather <- weather %>%
  mutate_if(is.character, list(~na_if(., "")))
lapply(weather, identifyMissing)

# remove values that don't fit data type
weather <- weather %>%
  mutate(Precipitation_Belfort = na_if(Precipitation_Belfort, ".33 avg 24-27th")) %>% 
  mutate(Precipitation_ETI = na_if(Precipitation_ETI, ".")) %>% 
  mutate(Precipitation_ETI = na_if(Precipitation_ETI, ".15.")) %>% 
  mutate(Precipitation_ETI = na_if(Precipitation_ETI, ".28.09")) %>% 
  mutate(Precipitation_ETI = na_if(Precipitation_ETI, "NA")) %>% 
  mutate(Snowfall = na_if(Snowfall, '.5"')) %>% 
  mutate(Snow_On_Ground = na_if(Snow_On_Ground, ".6.5")) %>% 
  mutate(Wind_Direction = na_if(Wind_Direction, "8")) %>% 
  mutate(Wind_Direction = na_if(Wind_Direction, "88")) %>% 
  mutate(Wind_Direction = na_if(Wind_Direction, "9")) %>% 
  mutate(Wind_Direction = na_if(Wind_Direction, "99")) %>% 
  mutate(Sky_Cover = na_if(Sky_Cover, "80")) %>% 
  mutate(Notes = na_if(Notes, "8"))
```

Fix types of all data fields
```{r}
# add all flag columns for numeric variable, note trace values in flag column
weather <- weather %>% 
  mutate(Flag_Max_Temperature = NA) %>% 
  mutate(Flag_Min_Temperature = NA) %>% 
  mutate(Flag_Current_Temperature = NA) %>% 
  mutate(Flag_Precipitation_Belfort = ifelse(Precipitation_Belfort == "T", "T", NA)) %>% 
  mutate(Flag_Precipitation_ETI = ifelse(Precipitation_ETI == "T", "T", NA)) %>% 
  mutate(Flag_Snowfall = ifelse(Snowfall == "T", "T", NA)) %>% 
  mutate(Flag_Snow_On_Ground = ifelse(Snow_On_Ground == "T", "T", NA)) %>% 
  mutate(Flag_Wind_Speed = NA) %>% 
  mutate(Flag_Evaporation_Pan = NA) %>% 
  mutate(Flag_Lake_Level = NA)

# change T values to NA
weather <- weather %>% 
  mutate(Precipitation_Belfort = na_if(Precipitation_Belfort, "T")) %>% 
  mutate(Precipitation_ETI = na_if(Precipitation_ETI, "T")) %>% 
  mutate(Snowfall = na_if(Snowfall, "T")) %>% 
  mutate(Snow_On_Ground = na_if(Snow_On_Ground, "T"))

# typecast columns that are not factors
weather <- weather %>% 
  mutate(Max_Temperature = as.integer(Max_Temperature)) %>% 
  mutate(Min_Temperature = as.integer(Min_Temperature)) %>% 
  mutate(Current_Temperature = as.integer(Current_Temperature)) %>% 
  mutate(Precipitation_Belfort = as.numeric(Precipitation_Belfort)) %>%
  mutate(Precipitation_ETI = as.numeric(Precipitation_ETI)) %>% 
  mutate(Snowfall = as.numeric(Snowfall)) %>% 
  mutate(Snow_On_Ground = as.numeric(Snow_On_Ground)) %>% 
  mutate(Wind_Speed = as.numeric(Wind_Speed)) %>% 
  mutate(Sky_Cover = as.integer(Sky_Cover)) %>% 
  mutate(Evaporation_Pan = as.numeric(Evaporation_Pan)) %>% 
  mutate(Lake_Level = as.numeric(Lake_Level))

lapply(weather, identifyMissing)
```

Fix factor types
```{r}
# convert degrees to cardinal direction
# degreeToDirection <- function(degree){
#   # if it's a number, convert
#   if (!is.na(as.numeric(degree))) {
#     d <- as.numeric(degree)
#     if ((d > 348.75 & d <= 360) | (d <= 11.25 & d >= 0)) {
#       return("N")
#     } else if (d > 11.25 & d <= 33.75) {
#       return("NNE")
#     } else if (d > 33.75 & d <= 56.25) {
#       return("NE")
#     } else if (d > 56.25 & d <= 78.75) {
#       return("ENE")
#     } else if (d > 78.75 & d <= 101.25) {
#       return("E")
#     } else if (d > 101.25 & d <= 123.75) {
#       return("ESE")
#     } else if (d > 123.75 & d <= 146.25) {
#       return("SE")
#     } else if (d > 146.25 & d <= 168.75) {
#       return("SSE")
#     } else if (d > 168.75 & d <= 191.25) {
#       return("S")
#     } else if (d > 191.25 & d <= 213.75) {
#       return("SSW")
#     } else if (d > 213.75 & d <= 236.25) {
#       return("SW")
#     } else if (d > 236.25 & d <= 258.75) {
#       return("WSW")
#     } else if (d > 258.75 & d <= 281.25) {
#       return("W")
#     } else if (d > 281.25 & d <= 303.75) {
#       return("WNW")
#     } else if (d > 303.75 & d <= 326.25) {
#       return("NW")
#     } else if (d > 326.25 & d <= 348.75) {
#       return("NNW")
#     }
#     return("NA")
#   } else { # if not a number, just return it
#     return(degree)
#   }
# }


# fix factors for Wind Direction
weather$Wind_Direction <- toupper(weather$Wind_Direction)

directions = c("N", "S", "E", "W", "NW", "NE", "SW", "SE", "NNW", "WNW", "NNE", "ENE", "WSW", "SSW", "ESE", "SSE")

# replace values that aren't in the valid options
mask <- weather$Wind_Direction %in% directions
weather <- weather %>% 
  mutate(Wind_Direction = replace(Wind_Direction, !mask, NA))

weather <- weather %>% 
  mutate(Wind_Direction = as.factor(Wind_Direction))

identifyLoners(weather$Wind_Direction)


# fix factors for cloud type
weather$Cloud_Type <- toupper(weather$Cloud_Type)

# check existing cloud types
weather %>% 
  group_by(Cloud_Type) %>%
  summarise(n_rows = length(Cloud_Type))

# replace values that aren't in the valid options

clouds = c("CI", "CC", "CS", "AC", "AS", "NS", "SC", "ST", "CU", "CB")

mask <- weather$Cloud_Type %in% clouds
weather <- weather %>%
  mutate(Cloud_Type = replace(Cloud_Type, !mask, NA))

weather <- weather %>% 
  mutate(Cloud_Type = as.factor(Cloud_Type))

identifyLoners(weather$Cloud_Type)

# replace abbreviations
levels(weather$Cloud_Type)
levels(weather$Cloud_Type)[1] <- "ALTOCUMULUS"
levels(weather$Cloud_Type)[2] <- "ALTOSTRATUS"
levels(weather$Cloud_Type)[3] <- "CIRROCUMULUS"
levels(weather$Cloud_Type)[4] <- "CIRRUS"
levels(weather$Cloud_Type)[5] <- "CIRROSTRATUS"
levels(weather$Cloud_Type)[6] <- "CUMULUS"
levels(weather$Cloud_Type)[7] <- "NIMBOSTRATUS"
levels(weather$Cloud_Type)[8] <- "STRATOCUMULUS"
levels(weather$Cloud_Type)[9] <- "STRATUS"

# check
weather %>% 
  group_by(Cloud_Type) %>%
  summarise(n_rows = length(Cloud_Type))
```

Check bounds on all data types
```{r}
# check min and max dates for out-of-bound dates
min(weather$Date)
max(weather$Date)

# flag out-of-bounds temperatures
weather <- weather %>% 
  mutate(Flag_Max_Temperature = ifelse((Max_Temperature > 120) | (Max_Temperature < -20), "R", NA)) %>% 
  mutate(Flag_Min_Temperature = ifelse((Min_Temperature > 120) | (Min_Temperature < -40), "R", NA)) %>%
  mutate(Flag_Current_Temperature = ifelse((Current_Temperature > 120) | (Current_Temperature < -40), "R", NA))

# replace out-if-bounds temperatures with historic temperatures taken from https://www.wunderground.com/history for 
weather %>% filter(Flag_Max_Temperature == "R")
weather$Max_Temperature[weather$Max_Temperature == 725] <- 82
weather$Max_Temperature[weather$Max_Temperature == 6261] <- 77

weather %>% filter(Flag_Min_Temperature == "R")
weather$Min_Temperature[weather$Min_Temperature == 7261] <- 72

weather %>% filter(Flag_Current_Temperature == "R")
weather$Min_Temperature[weather$Current_Temperature == 354] <- 35
weather$Min_Temperature[weather$Current_Temperature == 598] <- 65
weather$Min_Temperature[weather$Current_Temperature == 239] <- 37

# flag other out-of-bounds measurements
weather <- weather %>% 
  mutate(Flag_Wind_Speed = ifelse(Wind_Speed > 100, "R", NA))
    
mask <- weather$Precipitation_ETI > 10  # avoid overwriting Ts with logical slicing
weather <- weather %>% 
  mutate(Flag_Precipitation_ETI = replace(Flag_Precipitation_ETI, mask, "R"))

mask <- weather$Snow_On_Ground > 100
weather <- weather %>% 
  mutate(Flag_Snow_On_Ground = replace(Flag_Snow_On_Ground, mask, "R"))

# make Notes consistent, remove commas to avoid confusion in CSV file
weather$Notes <- toupper(weather$Notes)
weather$Notes <- gsub(",", ".", weather$Notes)

# flag possible instrumentation changes
mask <- grepl("UV-B", weather$Notes, ignore.case = TRUE)
weather <- weather %>% 
  mutate(Flag_Max_Temperature = replace(Flag_Max_Temperature, mask, "I")) %>% 
  mutate(Flag_Min_Temperature = replace(Flag_Min_Temperature, mask, "I")) %>% 
  mutate(Flag_Current_Temperature = replace(Flag_Current_Temperature, mask, "I"))

mask <- grepl("iButton", weather$Notes, ignore.case = TRUE) | grepl("thermochron", weather$Notes, ignore.case = TRUE)
weather <- weather %>% 
  mutate(Flag_Max_Temperature = replace(Flag_Max_Temperature, mask, "I")) %>% 
  mutate(Flag_Min_Temperature = replace(Flag_Min_Temperature, mask, "I")) %>% 
  mutate(Flag_Current_Temperature = replace(Flag_Current_Temperature, mask, "I"))

mask <- grepl("airport", weather$Notes, ignore.case = TRUE)
weather <- weather %>%
  mutate(Flag_Snowfall = replace(Flag_Snowfall, mask, "I")) %>%
  mutate(Flag_Snow_On_Ground = replace(Flag_Snow_On_Ground, mask, "I"))

mask <- grepl("ETI", weather$Notes, ignore.case = TRUE)
weather <- weather %>%
  mutate(Flag_Precipitation_ETI = replace(Flag_Precipitation_ETI, mask, "I"))

mask <- grepl("evap", weather$Notes, ignore.case = TRUE)
weather <- weather %>%
  mutate(Flag_Evaporation_Pan = replace(Flag_Evaporation_Pan, mask, "I"))

# flag possible instrument error
mask <- weather$Wind_Speed <= 0
weather <- weather %>% 
  mutate(Flag_Wind_Speed = replace(Flag_Wind_Speed, mask, "D"))

# check precision based on metadata
weather$Precipitation_Belfort <- round(weather$Precipitation_Belfort, digits = 2)
weather$Precipitation_ETI <- round(weather$Precipitation_ETI, digits = 2)
weather$Snowfall <- round(weather$Snowfall, digits = 2)
weather$Snow_On_Ground <- round(weather$Snow_On_Ground, digits = 1)
weather$Wind_Speed <- round(weather$Wind_Speed, digits = 6)
```


Export
```{r}
# omit evaporation by Adam's request
weather <- weather %>% 
  select(-Evaporation_Pan, -Flag_Evaporation_Pan, -Lake_Level, -Flag_Lake_Level)

# remove wind due to instrument inconsistency
weather <- weather %>% 
  select(-Wind_Direction, -Wind_Speed, -Flag_Wind_Speed)

# sort
weather <- weather %>% 
  arrange(Date)

write.csv(weather, file = "weather.csv", eol = "\r\n", row.names = FALSE)
```
